\Title

### 2023-02-20 Grimaldi-etal-BiolCybernetics.tex
Update on Overleaf.

Of interest is to define the receptive field of a postsynaptic neuron $\synapse^\postsynaddr =  \{(\presynaddr_\ranksyn, \postsynaddr_\ranksyn, \synapticweight_\ranksyn, \synapticdelay_\ranksyn) \| \postsynaddr_\ranksyn=\postsynaddr\}_{\ranksyn \in [1,\Nsyn]}  \subset \synapse$, or the emitting field of a presynaptic neuron $\synapse_\presynaddr = \{(\presynaddr_\ranksyn, \postsynaddr_\ranksyn, \synapticweight_\ranksyn, \synapticdelay_\ranksyn) \| \presynaddr_\ranksyn=\presynaddr\}_{\ranksyn \in [1,\Nsyn]} \subset \synapse$. As a consequence, a postsynaptic neuron $\postsynaddr$ receives an event stream which is multiplexed by the synapses of its receptive field. It results in a weighted event stream (\seeFig{izhikevich}-\textit{(Middle-Left}) for each  postsynaptic neuron $\postsynaddr$: 
%
\begin{equation}\label{eq:stream_b}
\event_\postsynaddr = \{(\presynaddr_\arank, \synapticweight_\arank, \timev_\arank+\synapticdelay_\ranksyn) \| \presynaddr_\arank = \presynaddr_\ranksyn \}_{\arank \in [1,\numevent], \ranksyn \in \synapse^\postsynaddr}
\end{equation}
%
In biology, this new stream of events is naturally ordered in time as events reach the soma of postsynaptic neurons. However, it should be properly reordered in simulations. Crucially, when postsynaptic neurons are activated on their soma by a specific spatio-temporal motif which is imprinted in the set of synapses, the discharge probability will increase, notably when these spikes converge on the soma in a synchronous manner while the activation function of the neurons of the HD-SNN can be selected among the whole range of spiking neuron response functions (\seeFig{izhikevich}). We defined the HD-SNN model in this subsection in all generality, and in the next subsection, we describe an implementation of our model adapted for the motion detection task. 
%
\subsection{Application of the HD-SNN model to the motion detection task}
%
Such event-based model can be adapted to computer vision tasks. In particular, from the perspective of simulating such event-based computations on standard CPU- or GPU-based computers and then using parallel computing, it is useful to transform this event-based representation into a dense representation. Indeed, by discretization of time, we may transform any event-based input from an event-based camera as a Boolean matrix $A \in \{0, 1 \}^{\Npol \times \Ntime \times \Nx \times \Ny}$, defined for all polarities $p$, times $t$ and space coordinates $x$ and $y$. The values are by definition equal to zero except at the occurrences of events: $\forall \arank \in [1,\numevent]$, $A(\polev_\arank, t_\arank, x_\arank, y_\arank)=1$. First, one can note that by using a discretization, the computational brick used in equation~\eqref{eq:kernel_b} corresponds to a temporal convolution which transforms the input $A$ using one kernel per postsynaptic neuron~\citep{grimaldi_learning_2022}. To benefit from invariance to location observed in images and exploited in Convolutional Neural Networks, one may further assume that synaptic motifs should be similar across different positions such that we may define a spatio-temporal convolution operator. %The characteristics of the kernels are shared among one class of neurons such that $\forall i,j \in [1, N_\class]^2, \; \kernel^i_\class = \kernel^j_\class$.

% Here, we will consider that heterogeneous delays are limited in range
In particular, the whole synaptic set can be represented as one kernel for each class $\class$ of the supervision task as the dense matrix $\kernel_\class$ of size $(\Npol, \Ktime, \Kx, \Ky)$, where  $\Ktime$ is the number of delays and $\Kx$ and $\Ky$ are the number of pixels in both spatial dimensions. To keep an analogy with the HD-SNN model, for neuron $\postsynaddr$ of position $(x_\postsynaddr, y_\postsynaddr)$ and class $\class$, $\kernel_\class$ gives the weights as a function of relative addresses and synaptic delays: $\forall {\delta_x \in [1, \Kx], \delta_y \in [1, \Ky], \delta_\timev \in [1, \Ktime], \polev \in [1, \Npol]}, \; \kernel_\class(\polev, \delta_x, \delta_y, \synapticdelay_\timev) = \synapticweight$. The connectivity of neuron $\postsynaddr$ is defined locally, around it position $(x_\postsynaddr, y_\postsynaddr)$. %Therefore, if we make the approximation that one neuron integrates only on the temporal window given by its variety of synaptic delays, the integration of the spike train as input can be formalized by a 3D convolution operation. The longest synaptic delay defines the depth of the kernel $\kernel^\postsynaddr$ and all possible delays associated to the different presynaptic addresses are represented. 
Using this dense representation, the processing of $A$ by the model can be written as:
%
\begin{equation}\label{eq:kernel_b}
\forall x, y, \timev, \; \;
\mathcal{C}_\class(x, y, \timev)
= \sum_{p, \delta_x, \delta_y, \synapticdelay} \kernel_\class(p, \delta_x, \delta_y, \synapticdelay) \cdot A(p, x - \delta_x, y - \delta_y, \timev-\synapticdelay)
\end{equation}
%
where $\delta_x$ and $\delta_y$ are the relative adresses of the synapses inside a kernel and $\delta_\timev$ is the synaptic delay. 
This shows that $\mathcal{C}_\class$ is the result of a spatio-temporal convolution of the dense representation of the event stream with the dense kernels formed by the set of synapses:  $\mathcal{C}_\class = \kernel_\class \ast A$ (\seeFig{model} for an illustration). Note that to remain within the framework of a causal calculation, the kernels are shifted in time such that only past information gives an answer at the present time. This well-known computation defines a differentiable measure which is very efficiently implemented for GPUs and which we will use for learning the classification of different motifs in the event stream. The same type of spatio-temporal filtering is used as a pre-processing stage of a pattern recognition algorithm~\citep{ghosh_spatiotemporal_2019}. Also,~\citet{sekikawa_constant_2018} developed an efficient 3D convolutional algorithm which implements a motion estimation task. By assuming a locally constant velocity, the authors assume the 3D kernel can be decomposed into a 2D kernel representing the shapes and a 3D kernel representing the velocity.

Here, we keep the analogy with spiking neurons, and we try to use as few heuristics as possible to observe the detection of spiking motifs using the spatio-temporal kernels. If so, this precise spatio-temporal patterns prove to be of interest for neural computations and one can infer that biological neurons make use of this information as well. 
% \note{plus de biblio sur la motion detection/segmentation ou optique flow avec les events? (reviewer 2)}
The 3D convolution represents the linear integration of the spike train as the linear input to the neuron. Then, the activation function of our model is a softmax function implementing a form of Multinomial Logistic Regression (MLR)~\citep{grimaldi_robust_2022}, in analogy to a spiking Winner-Takes-All network~\citep{nessler_bayesian_2013}. In our MLR model, a probability value for each class (i.e. each motion direction) is predicted for each position $x, y$ and time $\timev$ as a softmax function of the linear combination of the list of events. % on the basal dendrite of a neuron $\postsynaddr$ in association to a specific class. The linear combination can be defined by a set of synapses $\synapse^\postsynaddr$ as described in the HD-SNN model. 
Formally, using the kernels, it transforms the input raster plot into a probability with the following formula:
%
\begin{equation}\label{eq:mlr}
    \forall x, y, \timev, \forall \class \in [1, N_\class], \; \;
Pr(k=\class \; \vert \;  x, y, \timev) =
\frac {\exp  (\mathcal{C}_\class(x, y, \timev) +\bias_\class) }{Z(x, y, \timev)}
\end{equation} 
%
where $Z(x, y, \timev)=\sum_{\class \in [1, N_\class]} \exp  (\mathcal{C}_\class(x, y, \timev) +\bias_\class)$ is the partition function and  $\bias_\class$ is the bias linked to class $\class$. 
In particular, we expect that some specific motifs may become tightly synchronized as they reach the basal dendritic tree, leading to a high postsynaptic activity which makes it progressively more likely to generate an output spike. The spiking output of the model corresponds to an event with the highest probability class.

Once the general framework used to solve the ecological task described in section~\ref{sec:task} explained, we can add some heuristics based on neuro-scientific observations to constrain our model and its strategies to solve the task. Note that the general framework is the same as the one presented in~\cite{grimaldi_learning_2022} and that, apart from the more complex task to solve and a deeper analysis of the results, these novel heuristics, inspired by neuroscience, are the main methodological differences with that previous study. In our simulations, We set the size of the kernels to $(21, 17, 17)$ and define as many classes as the number of motions (directions and speeds): $N_\class = 8 \times 3$. To avoid introducing biases in the directions, we apply a circular mask on the spatial dimensions of the kernels. And because the synaptic delay is linked, by physical principles, to the distance between the presynaptic neuron and the postsynaptic one, we apply a linear mask on the temporal axis such that the connectivity depends on the delay: the lower the delay, the lower the circular mask radius (\seeFig{kernels} for an illustration). The second heuristics is based on the observation that biological neurons have a sparse spiking activity. From this observation, we set an output firing rate for the layer of spiking neurons in order to make a sparse selection of the events at which the model makes a decision. It results in the selection of the coefficients with the maximum likelihood value. Therefore, we will only select a given proportion of the most active cells, which we set here to $0.1\%$ of all output voxels \note{change accordingly to the model's parameters}. This threshold was found using cross-validation and future work should automate how this may be determined for various datasets. Given a stimulus input, the model thus yields a spike output in the postsynaptic address space.

\note{describe the heuristics inspired by neuroscience, remaining complex cells}
\note{describe how we compute accuracy}

>>>>

Of interest is to define the receptive field of a postsynaptic neuron $\synapse^\postsynaddr =  \{(\presynaddr_\ranksyn, \postsynaddr_\ranksyn, \synapticweight_\ranksyn, \synapticdelay_\ranksyn) \| \postsynaddr_\ranksyn=\postsynaddr\}_{\ranksyn \in [1,\Nsyn]}  \subset \synapse$, or the emitting field of a presynaptic neuron $\synapse_\presynaddr = \{(\presynaddr_\ranksyn, \postsynaddr_\ranksyn, \synapticweight_\ranksyn, \synapticdelay_\ranksyn) \| \presynaddr_\ranksyn=\presynaddr\}_{\ranksyn \in [1,\Nsyn]} \subset \synapse$. As a consequence, a postsynaptic neuron $\postsynaddr$ receives an event stream which is multiplexed by the synapses of its receptive field. It results in a weighted event stream (\seeFig{izhikevich}-\textit{(Middle-Left}) for each  postsynaptic neuron $\postsynaddr$: 
%
\begin{equation}\label{eq:stream_b}
\event_\postsynaddr = \{(\presynaddr_\arank, \synapticweight_\arank, \timev_\arank+\synapticdelay_\ranksyn) \| \presynaddr_\arank = \presynaddr_\ranksyn \}_{\arank \in [1,\numevent], \ranksyn \in \synapse^\postsynaddr}
\end{equation}
%
In biology, this new stream of events is naturally ordered in time as events reach the soma of postsynaptic neurons. However, it should be properly reordered in simulations. Crucially, when postsynaptic neurons are activated on their soma by a specific spatio-temporal motif which is imprinted in the set of synapses, the discharge probability will increase, notably when these spikes converge on the soma in a synchronous manner while the activation function of the neurons of the HD-SNN can be selected among the whole range of spiking neuron response functions (\seeFig{izhikevich}). We defined the HD-SNN model in this subsection in all generality, and in the next subsection, we describe an implementation of our model adapted for the motion detection task. 
%
\subsection{Application of the HD-SNN model to the motion detection task}
%
Indeed, such an event-based model can be easily adapted to common computer vision tasks. In particular, from the perspective of simulating such event-based computations on standard CPU- or GPU-based computers and then using parallel computing, it is useful to transform this event-based representation into a dense representation. Indeed, by discretization of time, we may transform any event-based input from an event-based camera as a Boolean matrix $A \in \{0, 1 \}^{\Npol \times \Ntime \times \Nx \times \Ny}$, defined for all polarities $p$, times $t$ and space coordinates $x$ and $y$. The values are by definition equal to zero except at the occurrences of events: $\forall \arank \in [1,\numevent]$, $A(\polev_\arank, t_\arank, x_\arank, y_\arank)=1$. First, one can note that by using a discretization, the computational brick used in equation~\eqref{eq:kernel_b} corresponds to a temporal convolution which transforms the input $A$ using one kernel per postsynaptic neuron~\citep{grimaldi_learning_2022}. To benefit from invariance to location observed in images and exploited in Convolutional Neural Networks, one may further assume that synaptic motifs should be similar across different positions such that we may define a spatio-temporal convolution operator. %The characteristics of the kernels are shared among one class of neurons such that $\forall i,j \in [1, N_\class]^2, \; \kernel^i_\class = \kernel^j_\class$.

In particular, the whole synaptic set can be represented as one kernel for each class $\class$ of the supervision task as the dense matrix $\kernel_\class$ of size $(\Npol, \Ktime, \Kx, \Ky)$, where  $\Ktime$ is the number of delays and $\Kx$ and $\Ky$ are the number of pixels in both spatial dimensions. To keep an analogy with the HD-SNN model, for neuron $\postsynaddr$ of position $(x_\postsynaddr, y_\postsynaddr)$ and class $\class$, $\kernel_\class$ gives the weights as a function of relative addresses and synaptic delays: $\forall {\delta_x \in [1, \Kx], \delta_y \in [1, \Ky], \delta_\timev \in [1, \Ktime], \polev \in [1, \Npol]}, \; \kernel_\class(\polev, \delta_x, \delta_y, \synapticdelay_\timev) = \synapticweight$. The connectivity of neuron $\postsynaddr$ is defined locally, around it position $(x_\postsynaddr, y_\postsynaddr)$. %Therefore, if we make the approximation that one neuron integrates only on the temporal window given by its variety of synaptic delays, the integration of the spike train as input can be formalized by a 3D convolution operation. The longest synaptic delay defines the depth of the kernel $\kernel^\postsynaddr$ and all possible delays associated to the different presynaptic addresses are represented. 
Using this dense representation, the processing of $A$ by the model can be written as:
%
\begin{equation}\label{eq:kernel_b}
\forall x, y, \timev, \; \;
\mathcal{C}_\class(x, y, \timev)
= \sum_{p, \delta_x, \delta_y, \synapticdelay} \kernel_\class(p, \delta_x, \delta_y, \synapticdelay) \cdot A(p, x - \delta_x, y - \delta_y, \timev-\synapticdelay)
\end{equation}
%
where $\delta_x$ and $\delta_y$ are the relative adresses of the synapses inside a kernel and $\delta_\timev$ is the synaptic delay. 
This shows that $\mathcal{C}_\class$ is the result of a spatio-temporal convolution of the dense representation of the event stream with the dense kernels formed by the set of synapses:  $\mathcal{C}_\class = \kernel_\class \ast A$ (\seeFig{model} for an illustration). Note that to remain within the framework of a causal calculation, the kernels are shifted in time such that only past information gives an answer at the present time. This well-known computation defines a differentiable measure which is very efficiently implemented for GPUs and which we will use for learning the classification of different motifs in the event stream. The same type of spatio-temporal filtering is used as a pre-processing stage of a pattern recognition algorithm~\citep{ghosh_spatiotemporal_2019}. Also,~\citet{sekikawa_constant_2018} developed an efficient 3D convolutional algorithm which implements a motion estimation task. By assuming a locally constant velocity, the authors assume the 3D kernel can be decomposed into a 2D kernel representing the shapes and a 3D kernel representing the velocity.

Here, we keep the analogy with spiking neurons and we try to observe the detection of spiking motifs using the spatio-temporal kernels. If so, this precise spatio-temporal patterns prove to be of interest for neural computations and one can infer that biological neurons make use of this information as well. 
% \note{plus de biblio sur la motion detection/segmentation ou optique flow avec les events? (reviewer 2)}
The 3D convolution represents the linear integration of the spike train as the linear input to the neuron. Then, the activation function of our model is a softmax function implementing a form of Multinomial Logistic Regression (MLR)~\citep{grimaldi_robust_2022}, in analogy to a spiking Winner-Takes-All network~\citep{nessler_bayesian_2013}. In our MLR model, a probability value for each class (i.e. each motion direction) is predicted for each position $x, y$ and time $\timev$ as a softmax function of the linear combination of the list of events. % on the basal dendrite of a neuron $\postsynaddr$ in association to a specific class. The linear combination can be defined by a set of synapses $\synapse^\postsynaddr$ as described in the HD-SNN model. 
Formally, using the kernels, it transforms the input raster plot into a probability with the following formula:
%
\begin{equation}\label{eq:mlr}
    \forall x, y, \timev, \forall \class \in [1, N_\class], \; \;
Pr(k=\class \; \vert \;  x, y, \timev) =
\frac {\exp  (\mathcal{C}_\class(x, y, \timev) +\bias_\class) }{Z(x, y, \timev)}
\end{equation} 
%
where $Z(x, y, \timev)=\sum_{\class \in [1, N_\class]} \exp  (\mathcal{C}_\class(x, y, \timev) +\bias_\class)$ is the partition function and  $\bias_\class$ is the bias linked to class $\class$. 
In particular, we expect that some specific motifs may become tightly synchronized as they reach the basal dendritic tree, leading to a high postsynaptic activity which makes it progressively more likely to generate an output spike. The spiking output of the model corresponds to an event with the highest probability class.

Now that this general framework explained, we can add some heuristics based on neuro-scientific observations to constrain our model and its strategies to solve the task  used to solve the ecological task described in section~\ref{sec:task}. Note that the general framework is the same as the one presented in~\cite{grimaldi_learning_2022} and that, apart from the more complex task to solve and a deeper analysis of the results, these novel heuristics, inspired by neuroscience, are the main methodological differences with that previous study. In our simulations, We set the size of the kernels to $(21, 17, 17)$ and define as many classes as the number of motions (directions and speeds): $N_\class = 8 \times 3$. To avoid introducing biases in the directions, we apply a circular mask on the spatial dimensions of the kernels. First, we observed that moving images produced trajectories of ON and OFF spikes, and that these were present in both arrangements of polarities. This is due to the fact that our whitened images show a symmetry in the luminance profiles, and that an image with inverted contrast is indistinguishable. As this arrangement of polarities is independent of the speed, we added a mechanism gathering the linear values for the movie and the movie with ON and OFF cells flipped, and kept only the maximal value for each voxel. This is similar to the computation operated for complex cells in the primary visual cortex.  Second, we included a prior in the motions that could be selected, as there is a prior in natural scenes for slow speeds~\citep{vacher_bayesian_2018}. As here we aim at capturing the possible convergence of events' trajectories converging to each voxel, we thus apply a mask on the spatio-temporal kernels such that the connectivity is bounded by the delay: the lower the delay, the lower the circular mask radius (\seeFig{kernels} for an illustration). This is compatible with neuroscientific principles because of the limits in conduction delay along horizontal connections, the synaptic delay is linked, by physical principles, to the distance between the presynaptic neuron and the postsynaptic one.

% The second heuristics is based on the observation that biological neurons have a sparse spiking activity.  From this observation, we set an output firing rate for the layer of spiking neurons in order to make a sparse selection of the events at which the model makes a decision. It results in the selection of the coefficients with the maximum likelihood value. Therefore, we will only select a given proportion of the most active cells, which we set here to $0.1\%$ of all output voxels \note{change accordingly to the model's parameters}. This threshold was found using cross-validation and future work should automate how this may be determined for various datasets. Given a stimulus input, the model thus yields a spike output in the postsynaptic address space.
% \note{describe the heuristics inspired by neuroscience, remaining complex cells}
\note{describe how we compute accuracy}
